A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-02 linear.tex
more expansion on dust-drag numerical section

\subsection{Overstable epicycles}
If we further consider the limit $k_z=0$, the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}, reduces to 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\tepsilon \taus}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
Here we use $\tau_\mathrm{s}=\tstop/(1-\tepsilon)$ for convenience. 
If the quadratic term is neglected, then $\imag{\sigma}<0$, implying
stability, as concluded by \cite{youdin05a}.

However, retaining all the terms (but still in the incompressible
limit) can yield instability, as follows. Eq. \ref{kz_zero_si} admit
pure oscillations with $\sigma = -\omega$ when 
\begin{align*}
  \omega &= \pm \kappa, \\
  \taus & = \frac{\epsilon\omega}{\tepsilon k _x F_r }
         \equiv \tau_{\mathrm{s}0}.
\end{align*} 
Since $\taus\geq 0$, we require $k_xF_r > 0$ if $\omega = \kappa$;
and $k_xF_r < 0$ if $\omega = -\kappa$. For $k_x>0$ these marginally
stable epicycles require radial pressure gradients decreasing and increasing
outwards, respectively.

@@

We now perturb about this state of marginal stability by writing
$\sigma = - \omega + \dd \sigma$ with $\dd\sigma = \ii\dd s -
\dd\omega$, $\taus = \tau_{\mathrm{s}0} +
\dd\taus$, and assume the $\dd$ quantities are small in
magnitude. Then 
\begin{align*} 
\dd s = \frac{2\tepsilon\kappa^2 }{1 +
  4\tepsilon^2\tau_{\mathrm{s}0}^2\kappa^2}\dd \taus. 
\end{align*}
Thus given $\tepsilon$ and $k_x$, there exist growing modes
when the stopping time is slightly larger than
$\tau_{\mathrm{s}c}$. These correspond to slowly-growing epicycles
\citep[cf. the `convective overstability', ][]{klahr14,lyra14,latter16}.

@@

@@

We now consider the streaming instability with $k_z\neq 0$. 
We solve Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$.

@@

>>>>

\subsection{Overstable epicycles}
If we further consider the limit $k_z=0$, the incompressible dispersion
relation, Eq. \ref{streaming_incompressible}, reduces to 
\begin{align}      %streaming_incompressible
\ii \sigma^2 - \frac{\sigma}{\epsilon \tstop}  - \left(\ii \kappa^2 +
  \frac{k_xF_r}{\epsilon}\right) = 0. \label{kz_zero_si}
\end{align} 
%Here we use $\tau_\mathrm{s}=\tstop/(1-\tepsilon)$ for convenience. 
If the quadratic term is neglected, then $\imag{\sigma}<0$, implying
stability, as concluded by \cite{youdin05a}.

However, retaining all the terms (but still in the incompressible
limit) can yield instability, as follows. Eq. \ref{kz_zero_si} admit
pure oscillations with $\sigma = -\omega$ when 
\begin{align*}
  \omega &= \pm \kappa, \\
  \tstop & = \frac{\omega}{ k _x F_r }
         \equiv t_{\mathrm{s}0}.
\end{align*} 
Since $\tstop\geq 0$, we require $k_xF_r > 0$ if $\omega = \kappa$;
and $k_xF_r < 0$ if $\omega = -\kappa$. For $k_x>0$ these marginally
stable epicycles require radial pressure gradients decreasing and increasing
outwards, respectively.

@@

We now perturb about this state of marginal stability by writing
$\sigma = \dd \sigma - \omega $ with $\dd\sigma = \ii\dd s -
\dd\omega$, $\tstop = t_{\mathrm{s}0} +
\dd\tstop $, and assume the $\dd$ quantities are small in
magnitude. Then 
\begin{align*} 
\dd s = \frac{2\epsilon \kappa^2 }{1 +
  4\epsilon^2 t_{\mathrm{s}0}^2\kappa^2}\dd \tstop. 
\end{align*}
Thus given $k_xF_r$, there exist growing modes 
when the stopping time is slightly larger than
$t_{\mathrm{s}0}$.

@@

If we instead fix $\tstop$ and vary $k_xF_r$ about marginal stability,
a similar exercise yields  
\begin{align*} 
  \dd s = \frac{\epsilon \tstop^3}{1 +
  4\epsilon^2 \tstop^2 \kappa^2}\dd \left(k_x^2F_r^2\right).  
\end{align*}
Thus it is possible to destabilize the system by slightly increasing
$|k_xF_r|>\kappa/ \tstop$ (i.e. shorter radial wavelengths and/or
stronger pressure gradients).

\citep[cf. the `convective overstability', ][]{klahr14,lyra14,latter16}.

@@

We now consider the streaming instability with $k_z\neq 0$. 
We solve Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

@@

\subsection{Numerical example}

To demonstrate the above instabilities we numerically solve the
full dispersion relation, Eq. \ref{streaming_dispersion} at fixed $K_x
= 1500$ and $\tepsilon=2$. Growth rates are shown in 
Fig. \ref{dusty_growth} as a function of $K_z$ at fixed $\taus\OmK=0.01$
(top panel, which includes `linC' in Table \ref{si_compare}); and as a
function of $\taus\OmK$ at fixed $K_z=50$ (bottom panel).

The top panel shows that the streaming instability vanishes as
$K_z\to0$, but it dominates most of $K_z$
-space. The overstable epicycles become important for 
$K_z\lesssim 100$, but their growth rates are weak. We thus
expect streaming instabilities to dominate, provided small
lengthscales can be resolved.

In the bottom panel we confirm that the overstable epicycles require a
sufficiently large stopping time at fixed wavenumbers. The vertical
dashed line is the critical stopping time 
\begin{align*}
  \tau_{\mathrm{s}0}\OmK = \frac{1}{2K_x\left(1-\tepsilon\right)^2} 
\end{align*}
in the above notation. For this example, $\tau_{\mathrm{s}0}\OmK\simeq
0.003$, as observed. On the other hand, the streaming instability
shows weak dependence on $\taus$ for this range of stopping times.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/streaming2}
  \includegraphics[width=\linewidth]{figures/streaming3}
  \caption{Growth rates of dust-drag instabilities with fixed radial
    wavenumber and dust-to-gas ratio, as a function of
    the vertical wavenumber at fix stopping time (top); and as a
    function of stopping time at fixed vertical wavenumber (bottom).  
\label{dusty_growth}
  }
\end{figure}
